%&subfile
\begin{document}
\frontmatter
\title{Brief lectures notes on Stochastic Calculus for Finance}
\author{Gautam Iyer}
\email{gautam@math.cmu.edu}
\date{Spring, 2017}
\hypersetup{pageanchor=false}
\maketitle%

This work is licensed under the \emph{Creative Commons Attribution - Non Commercial - Share Alike 4.0 International License.}
This means you may adapt and or redistribute this document \textbf{for non commercial purposes}, provided you give appropriate credit and re-distribute your work under the same licence.
To view the full terms of this license, visit \url{http://creativecommons.org/licenses/by-nc-sa/4.0/} or send a letter to Creative Commons, PO Box 1866, Mountain View, CA 94042, USA.
\maketitle

\chapter*{Preface}

These are extremely brief notes written while teaching \emph{46-944: Stochastic Calculus for Finance I}.
The notes are extremely minimal and are mainly a ``better handwriting'' version of what I write in class.
I will usually also provide verbal explanation and intuition in class, which you \replace{will}{might} not find here.
Consequently, these notes will only help you if you are attending class, reading the material from elsewhere, or already know the material.

For more complete references, I recommend the following:
\begin{enumerate}
  \item
    \emph{Stochastic Calculus for Finance II} by Steven Shreve.
  \item 
    \emph{\href{http://bass.math.uconn.edu/finlmath.pdf}{The basics of Financial Mathematics}} by Rich Bass.
  \item
    \emph{Introduction to Stochastic Calculus with Applications} by Fima C Klebaner.
\end{enumerate}

\bigskip

\begin{note}[Contributing]
  These notes are open-source.
  If you want to fill in a few details, or add a section on something related \emph{I welcome your help}.
  The \LaTeX\ source for these notes is currently publicly hosted at GitLab: \url{https://gitlab.com/gi1242/cmu-mscf-944}.
  Please contact me (via email) if you are interested in contributing.
\end{note}

%\setcounter{tocdepth}{2}
\tableofcontents\thispagestyle{empty}
\mainmatter
\hypersetup{pageanchor=true}
\end{document}
